Nonparametric Clustering from Energy Statistics

### 2017-06-26 qcqp_clustering/draft/energy_clustering.tex
first finished draft

Mention why energy is important, main results, where it was applied, etc.
Motivate how this can be used for clustering. Mention most important
papers on this \ldots Explain main results of this paper and give a brief
outline.

@@

Before proposing a more general iterative algorithm to \eqref{eq:qcqp2},
let us consider two simple experiments with equal number of points
in each cluster. 
We keep increasing the number of points in the clusters for each experiment, 
and cluster the data using Algorithm~\ref{algo1d}. 
We also cluster the same data set
with GMM, through EM algorithm, and with $k$-means. In both of these
cases we use the initialization from $k$-means++ \cite{Vassilvitskii} 
and we run the algorithms
few times with different initializations and choose the answer
with best objective function value. We use \eqref{eq:accuracy} to measure
the clustering quality. 
In  Fig.~\ref{fig:1d}a 
we have data from normal distributions,
where we can see that all the three methods
perform closely, with a slight advantage of GMM, as expected, since
it is the right model for the data. However, as shown in Fig~\ref{fig:1d}b,
for lognormal distributions, Algorithm~\ref{algo1d} provides a huge improvement
compared to both GMM and $k$-means which basically cluster at chance.
The zero accuracy values for GMM happened when EM algorithm was unable
to estimate the parameters. These two simple experiments illustrate
how energy statistics based clustering is nonparametric, since it is able
to provide high quality clustering in settings where data comes
from very different distribtuions.

@@

In this section we will introduce a new iterative algorithm to find a local
maximizer of the QCQP \eqref{eq:qcqp2}, however, due to 
Proposition~\ref{th:kernel_kmeans} we can also find an approximate
solution by the well-known kernel $k$-means algorithm, which 
for convenience
will also be reestated in the present context.
First, let us introduce some base notation.

@@

In the experiments below we fix the semimetric 
according to the traditional energy distance \eqref{eq:energy}, and
the fixed point $x_0=0$ is chosen in the associated kernel  
\eqref{eq:kernel_semimetric}:
\begin{equation}
\rho(x,y) = \| x-y\|, \qquad \kk(x,y) = 
\tfrac{1}{2}\left( \| x \| + \| y \| - \| x-y \| \right).
\end{equation}
We may consider other semimetrics/kernels as well, 
but the above will be considered the
standard kernel for energy statistics and will always be present in every
experiment. Another usual choice would be
to use $\alpha\ne 1$ according to \eqref{eq:energy2}.

We will compare Algorithm~\ref{algo} with
kernel $k$-means Algorithm~\ref{kmeans_algo}, and also with
standard $k$-means and GMM through the expectation maximization (EM) 
algorithm
as well. Since for synthetic data the true labels are available,
our measure of clustering quality
will be \eqref{eq:accuracy}. Moreover, for all algorithms, we always
choose the initialization from $k$-means++ \cite{Vassilvitskii}.

We first consider clustering in high dimensions and analyze 
how the algorithms degrade as the number of dimensions increase, while
keeping the number of points in each cluster fixed. 
In Figure~\ref{fig:gauss}a we have data generated from normal 
distributions in 
$D$-dimensions:
\begin{equation}
\label{eq:gauss1}
\begin{split}
&x \sim \tfrac{1}{2}\left[ 
\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
For each experiment we only keep signal in
$\mu_2$ in the first $10$
dimensions, and keep increasing the ambient dimension $D$. For each
data set, we run the algorithms $10$ times and pick the result with
the best objective function value.
In Figure~\ref{fig:gauss}a one sees that
GMM is not able to estimate the covariance matrix 
when the number
of dimensions exceeds the number of points in each cluster. 
One can see a slightly better performance of Algorithm~\ref{algo} compared
to the other ones.
In Figure~\ref{fig:gauss}b we have the same type of experiment but 
with 
\begin{equation}
\label{eq:gauss2}
\begin{split}
&x \sim \tfrac{1}{2}\left[ 
\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \,
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \,
\Sigma_1 = I_D, \, 
\Sigma_2 = \left( \begin{smallmatrix} \tfrac{1}{2} I_{10} & 0 \\ 0 & I_{D-10}
\end{smallmatrix}\right). \quad
\end{split}
\end{equation}

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-.8em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss_cov.pdf}\\[-.8em]
(b)
\end{minipage}
\caption{
\label{fig:gauss}
Two clusters with normally distributed data, with $100$ points in each
cluster. We analyse the effect of ambient dimension while keeping Bayes
error fixed.
(a) 
We increase $D$ as described in \eqref{eq:gauss1}. The blue
line correspond to Algorithm~\ref{algo}, while
the magenta line 
corresponds to Algorithm~\ref{kmeans_algo}.
(b) The same but with data following \eqref{eq:gauss2}.
In both experiments one notice a slightly better accuracy
of Algorithm~\ref{algo} compared to the other ones.
}
\end{figure}

In Figure~\ref{fig:unbalanced}a we consider how the previous algorithms
behave
for unbalanced clusters. We generate data as
\begin{equation}
\label{eq:gauss3}
\begin{split}
&x \sim  
\dfrac{n_1}{N} \mathcal{N}(\mu_1,\Sigma_1) + 
\dfrac{n_2}{N} \mathcal{N}(\mu_2, \Sigma_2), 
\quad \mu_1 = (0,0,0,0)^\top , \,
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
&
\Sigma_1 = I_4, \quad
\Sigma_2 = \left( 
\begin{smallmatrix} 
1/2 & 0 & 0 & 0\\
0 & 1/2 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 
\end{smallmatrix}\right), \quad
n_1 = N - m, \quad  n_2 = N + m, \quad N=200,
\end{split}
\end{equation}
and we increase $m$, i.e. we make the clusters progressively more unbalanced.
As well-known, one an see that GMM
works well for unbalanced clusters, mostly because 
it is a soft clustering algorithm.
We can see that all the other kernel based methods
degrade more rapidly than GMM for
highly unbalanced clusters.
This is
unsurprising since Algorithms~\ref{algo}~and~\ref{kmeans_algo} both 
make hard assignments. Finally, Figure~\ref{fig:unbalanced}b considers
the same experiment as in Figure~\ref{fig:gauss}a but for lognormal
data. With $\rho(x,y) = \| x-y\|^{1/2}$ we even gain a slight improvement.
Energy statistics clustering outperforms all the other methods since
it is nonparametric.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss_pi.pdf}\\[-.8em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{loggauss_dim.pdf}\\[-.8em]
(b)
\end{minipage}
\caption{
\label{fig:unbalanced}
(a) Previous algorithms for unbalanced clusters, 
according to \eqref{eq:gauss3}.
(b) The same experiment as in Figure~\ref{fig:gauss}a but for lognormal
data as 
$x\sim \tfrac{1}{2}\left( e^{\mathcal{N}(\mu_1,\Sigma_1)} +
e^{\mathcal{N}(\mu_2,\Sigma_2)} \right)$ with
$\mu_1 = (0,\dotsc,0)^\top$,
$\mu_2 = 0.5\times(1,\dotsc,1,0,\dotsc,0)^\top$ with signal in $d=10$, 
$\Sigma_1 = 0.3 \times I_D$, and $\Sigma_2=I_D$. 
One see a clear advantage of
energy statistics clustering. The extra green line correspond
to $\rho(x,y) = \| x-y\|^{1/2}$.
}
\end{figure}

In Figure~\ref{fig:weird} we consider other types of data in two dimensions.
In each case we perform 10 experiments and compute the accuracy
based on the true labels. The plots show the average results of 10 runs
of each experiment (dots), and the maximum and minimum values as well (shaded
area). Although energy clustering is nonparametric, these datasets are
very specific so we can take advantage and make a suitable choice of
the semimetric $\rho$. This illustrates that energy clustering is also
flexible enough to incorporate prior information about the data.
In Figure~\ref{fig:weird}a we have parallel cigars from
\begin{equation}
\label{eq:cigar}
\begin{split}
&x \sim \tfrac{1}{2}\left( 
\mathcal{N}(\mu_1, \Sigma_1) + 
\mathcal{N}(\mu_2, \Sigma_2)
\right), \\
& \mu_1 = (0,0)^\top, \quad \mu_2 = (6.5, 0)^\top, \quad
\Sigma_1 = \Sigma_2 = \left( \begin{smallmatrix} 
1 & 0 \\ 0 & 20
\end{smallmatrix}\right)
\end{split}
\end{equation}
with $100$ points in each cluster. Notice that $k$-means is unable to
cluster this data, while energy clustering performs pretty much as well
as GMM, which is the optimal algorithm suited to this kind of data.
We used the semimetric $\rho(x,y) = \| x -  y\|^{1/2}$ in this case.
In Figure~\ref{fig:weird}b we generate two concentric circles with a small
gaussian noise. To make the semimetric a little bit more localized we
include a decaying exponential in the form:
\begin{equation}
\label{eq:rho_loc}
\rho(x,y) = \| x-y \|^2 e^{-\tfrac{1}{2}\| x-y \|}.
\end{equation}
Notice that energy clustering is able to cluster this data almost perfectly
in most cases.
In Figure~\ref{fig:weird}c we consider two spirals, also with a small
gaussian noise. Due to the geometry of the data
we consider
\begin{equation}
\label{eq:rho_sin}
\rho(x,y) = \| x - y\|^2 \theta \sin \left(  \tfrac{\| x-y \|}{\theta}\right),
\qquad \theta = 0.9.
\end{equation}
Again, energy clustering performs much better than $k$-means and GMM which
are unable to cluster this kind of dataset.

\begin{figure}
\begin{minipage}{0.33\textwidth}
\hspace{.7cm}\includegraphics[width=.82\textwidth]{cigar_data.pdf}
\end{minipage}
\begin{minipage}{0.33\textwidth}
\hspace{.7cm}\includegraphics[width=.82\textwidth]{circles_data.pdf}
\end{minipage}
\begin{minipage}{0.33\textwidth}
\hspace{.7cm}\includegraphics[width=.82\textwidth]{spirals_data.pdf}
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{cigars.pdf}\\[-.8em]
(a)
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{circles.pdf}\\[-.8em]
(b)
\end{minipage}
\begin{minipage}{0.33\textwidth}
\centering
\includegraphics[width=\textwidth]{spirals.pdf}\\[-.8em]
(c)
\end{minipage}
\caption{
\label{fig:weird}
(a) Parallel cigars generated from \eqref{eq:cigar} with $100$ points
in each class. Use use energy clustering
with $\rho = \| x - y\|^{1/2}$. (b) Two concentric circles with $150$ points
in each class. We use the semimetric \eqref{eq:rho_loc}. 
(c) Two concentric spirals with
$150$ points in each class. We use the semimetric \eqref{eq:rho_sin}.
}
\end{figure}

\section{Conclusion}

\subsection*{Acknowledgements}
We thank \ldots

>>>>

Energy statistics is based on the energy distance between probability
distributions, which provides a notion of statistical potential energy
to statistical observations, in close analogy to Newton's gravitational
potential energy in classical mechanics. We refer the reader to
\cite{Szkely2013}, and references therein, for an overview.
It has been applied to several goodness-of-fit hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, and (nonlinear)
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient. 
Moreover, energy statistics  was applied
to hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance.

More recently, the distance covariance was 
generalized from Euclidean
spaces to metric spaces of strong negative type \cite{Lyons}.
Furthermore, a unifying framework establishing an equivalence
between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of
distributions in reproducing kernel Hilbert spaces (RKHS), was
provided \cite{Sejdinovic2013}. This important work establishes the link
between techniques commonly used in the statistics literature, regarding
energy statistics, and techniques
commonly used in machine learning.

Given a dataset $\mathbb{X} = \{x_1, x_2,\dotsc,x_n \}$, where each
datapoint lives in a generic space $x_i \in \mathcal{X}$, the $k$-clustering
problem consists in grouping these points into $k$ groups $\{
\C_1,\C_2,\dotsc,\C_k \}$, such that
points belonging to the same group are more ``similar'' to each other
than to points in  other groups. This intuitive notion already assumes
a ``metric'' able to measure the similarity between datapoints. Clustering
is an unsupervised method, and one of the most important problems in machine
learning since it provides the first step towards automatically constructing
labels for previously unseen data. The most used algorithms are
by far $k$-means and gaussian mixture models (GMM) through expectation
maximization (EM) algorithm. Both are parametric, making 
strong assumptions about
the distribution of the data (normality), and the 
space where data lie is Euclidean, $\mathcal{X} = \mathbb{R}^D$, hence
the similarity is based on Euclidean metric, 
$\| x_i - x_j \|^2 = (x_i - x_j)^\top(x_i-x_j) = 
\sum_{\ell=1}^D (x_{i,\ell} - x_{j,\ell})^2$. These algorithms provide
a good clustering when data is linearly separable in Euclidean space.

To account for nonlinearly separable data, which may possibly live
in an arbitrary non-Euclidean space, kernel methods are usually employed.
If a Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$ 
is available \cite{Mercer}, which is a symmetric and positive definite 
function, it guarantees that there exists a map $\varphi : \mathcal{X} \to
\mathcal{H}$, where $\mathcal{H}$ is a Hilbert space. Thus, one can
compute similarities through the inner product and 
perform clustering in the feature space $\mathcal{H}$. The nice thing
about this is that the map $\varphi$ is only used implicitly, i.e. one
does not need to know $\varphi$, since the inner
product can be computed only based on the kernel $\kk(\cdot, \cdot)$, 
a convenient technique known as kernel trick. The well-known 
kernel $k$-means problem 
is exactly $k$-means in the feature space
\cite{Smola,Girolami} (see also \cite{Filippone} for a survey of clustering
methods). 
It exploits the nonlinear structure provided by
a kernel function to perform clustering in datasets that are not linearly
separable in their original space $\mathcal{X}$.
As it stands, kernel $k$-means is an heuristic approach in the sense that
it is not derived from a statistical theory. Moreover, the choice of
kernel is crucial, and there is no systematic theory for this. One
must rely on ad hoc methods.

In this paper, we consider how to perform clustering based on energy
statistics theory. The original goal is to provide a nonparametric
clustering method, since energy distance is nonparametric.
The basis of our work is the theory developed in 
\cite{Sejdinovic2013}. Based on this and on the multi-sample test
for equality of distributions from energy statistics, we show that
there is only one function that provides the clustering optimization
problem. We then show that this clustering
optimization problem reduces to a quadratically constrained quadratic
problem (QCQP). This optimization problem has the same form as kernel $k$-means,
spectral clustering, and some well-known graph partitioning problems
\cite{Dhillon}. This lead us to show that energy statistics based
clustering is actually equivalent to kernel $k$-means optimization problem.
However, energy statistics fixes a kernel to begin with.
Furthermore, these results can be seen as a first-principle derivation
of kernel $k$-means from energy statistics, which thus bring
these kernel based clustering methods into a unifying mathematical theory.
We also provide an iterative algorithm, different but with the same
complexity as kernel $k$-means algorithm. The numerical results indicate
that this method is more accurate and robust.

Our work is organized as follows. In section~\ref{sec:background} we review
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main results of this paper
where we consider a clustering theory based on energy statistics, leading
to a QCQP which is NP-hard.
In section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM algorithms.
Our conclusions are in section~\ref{sec:conclusion}.

@@

Before proposing a more general iterative algorithm to \eqref{eq:qcqp2},
let us consider two simple experiments with equal number of points
in each cluster. 
We keep increasing the number of points in the clusters for each experiment, 
and cluster the data using Algorithm~\ref{algo1d}. 
We also cluster the same data set
with GMM, through EM algorithm, and with $k$-means. In both of these
cases we use the initialization from $k$-means++ \cite{Vassilvitskii} 
and we run the algorithms
few times with different initializations and choose the answer
with best objective function value. We use \eqref{eq:accuracy} to measure
the clustering quality. 
In  Fig.~\ref{fig:1d}a 
we have data from normal distributions,
where we can see that all the three methods
perform closely, with a slight advantage of GMM, as expected, since
it is the right model for the data. However, as shown in Fig~\ref{fig:1d}b,
for lognormal distributions, Algorithm~\ref{algo1d} provides a huge improvement
compared to both GMM and $k$-means which basically cluster at chance.
The zero accuracy values for GMM happened when EM algorithm was unable
to estimate the parameters. These two simple experiments illustrate
how energy statistics based clustering is nonparametric, since it is able
to provide high quality clustering in settings where data comes
from very different distributions.

@@

In this section we will introduce a new iterative algorithm to find a local
maximizer of the QCQP \eqref{eq:qcqp2}, however, due to 
Proposition~\ref{th:kernel_kmeans} we can also find an approximate
solution by the well-known kernel $k$-means algorithm, which 
for convenience
will also be restated in the present context.
First, let us introduce some base notation.

@@

In the experiments below we fix the semimetric 
according to the traditional energy distance \eqref{eq:energy}, and
the point $x_0=0$ is chosen in the associated kernel  
\eqref{eq:kernel_semimetric}. We thus have
\begin{equation}
\label{eq:standard_metric}
\rho(x,y) = \| x-y\|, \qquad \kk(x,y) = 
\tfrac{1}{2}\left( \| x \| + \| y \| - \| x-y \| \right).
\end{equation}
We will consider other semimetrics/kernels as well, 
but the above will be considered the
standard kernel for energy statistics and will always be present in every
experiment as a reference. Notice that this is a convention, 
we could have chosen
any other semimetric as the standard.
One of the main goals of the following 
experiments is to compare Algorithm~\ref{algo} to
kernel $k$-means algorithm, described in Algorithm~\ref{kmeans_algo}. Thus,
for every kernel used in Algorithm~\ref{algo}, we also use the same 
kernel in Algorithm~\ref{kmeans_algo}. Another goal is to compare
Algorithm~\ref{algo} with $k$-means and GMM (through expectation maximization
algorithm), as these are the most used
clustering algorithms in practice.
Since for synthetic data the true labels are available,
our measure of clustering quality
will be \eqref{eq:accuracy}. Moreover, for all algorithms, we always
choose the initialization from $k$-means++ \cite{Vassilvitskii}.

We first consider clustering in high dimensions and analyze 
how the algorithms degrade as the number of dimensions increase, while
keeping the number of points in each cluster fixed. The Bayes error
is also kept fixed as ambient dimension increases.
In Figure~\ref{fig:gauss}a we have data generated from $D$-variate normal 
distributions as follows: 
\begin{equation}
\label{eq:gauss1}
\begin{split}
&x \sim \tfrac{1}{2}\left[ 
\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
We only keep signal in
in the first $10$
dimensions of $\mu_2$, and keep increasing the ambient dimension $D$. For each
$D$, we perform $100$ experiments, obtaining the clustering accuracy
for each algorithm.
We can see that
GMM is not able to estimate the covariance matrix 
when the number
of dimensions exceeds the number of points in each cluster, so it gives
zero accuracy for $D \gtrsim 100$.
In Figure~\ref{fig:gauss}b we have the same type of experiment but 
with 
\begin{equation}
\label{eq:gauss2}
\begin{split}
&x \sim \tfrac{1}{2}\left[ 
\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \,
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \,
\Sigma_1 = I_D, \, 
\Sigma_2 = \left( \begin{smallmatrix} \tfrac{1}{2} I_{10} & 0 \\ 0 & I_{D-10}
\end{smallmatrix}\right). \quad
\end{split}
\end{equation}
Therefore, for both experiments shown in Figure~\ref{fig:gauss}
we can see a better performance of Algorithm~\ref{algo} compared
to the other ones, in particular compared to kernel $k$-means algorithm,
where we recall that both
aim at optimizing the same problem (see Proposition~\ref{th:kernel_kmeans}). 
Also, notice that $k$-means
and GMM are consistently the right model for this dataset, so it is hard
to perform better than these algorithms in this current setting.
Notice that Algorithm~\ref{algo} is more robust as the ambient
dimension increases.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss_cov.pdf}\\[-1.0em]
(b)
\end{minipage}
\caption{
\label{fig:gauss}
Effect of increasing the
ambient dimension while keeping Bayes error fixed, for 
two clusters with normally distributed data with $100$ points in each
cluster. 
(a) 
We increase $D$ as described in \eqref{eq:gauss1}. The blue
line correspond to Algorithm~\ref{algo}, while
the magenta line 
corresponds to kernel $k$-means, Algorithm~\ref{kmeans_algo}.
(b) The same but with data following \eqref{eq:gauss2}.
One notices that
Algorithm~\ref{algo} is more robust than the other ones.
}
\end{figure}

In Figure~\ref{fig:unbalanced} we consider the effect of having 
unbalanced clusters. We generate data as
\begin{equation}
\label{eq:gauss3}
\begin{split}
&x \sim  
\dfrac{n_1}{N} \mathcal{N}(\mu_1,\Sigma_1) + 
\dfrac{n_2}{N} \mathcal{N}(\mu_2, \Sigma_2), 
\quad \mu_1 = (0,0,0,0)^\top , \,
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
&
\Sigma_1 = I_4, \quad
\Sigma_2 = \left( 
\begin{smallmatrix} 
1/2 & 0 & 0 & 0\\
0 & 1/2 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 
\end{smallmatrix}\right), \quad
n_1 = N - m, \quad  n_2 = N + m, \quad N=200.
\end{split}
\end{equation}
We then increase $m$, i.e. we make the clusters progressively more unbalanced.
We generate $100$ experiments for each $m$, and plot the clustering accuracy
versus $m$.
As expected, GMM
works better than the other algorithms in the case of unbalanced clusters. 
This is mostly due to its soft assignments.
We can see that the other methods based on hard assignments degrade similarly,
and  more rapidly than GMM. This indicates that a fuzzy version of
energy statistics clustering should compensate for this effect.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{gauss_pi.pdf}
\caption{
\label{fig:unbalanced}
Previous algorithms for unbalanced clusters, 
according to \eqref{eq:gauss3}.}
\end{figure}

Now, besides \eqref{eq:standard_metric} we consider two other semimetrics:
\begin{align}
\rho_{1/2}(x,y) &= \| x-y \|^{1/2}, & \kk(x,y) &= \tfrac{1}{2} \left( 
\| x \|^{1/2} + \| y \|^{1/2} 
- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
\rho_{e}(x,y) &= 2 - 2 e^{-\| x- y\|/2}, & \kk(x,y) &= e^{-\| x-y\|/2}.
\label{eq:rhoe}
\end{align}
In Figure~\ref{fig:consist}a we have data in $20$ dimensions distributed as
\begin{equation}
\label{eq:20gauss}
\begin{split}
x &\sim \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2,
\Sigma_2) \right], \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
\Sigma_2 = I_{20}.
\end{split}
\end{equation}
We increase the number of points in each cluster and show the clustering
accuracy with different algorithms. The new semimetrics \eqref{eq:rhohalf}
and $\eqref{eq:rhoe}$ are indicated in the legend. One can see that
Algorithm~\ref{algo} performs better than all the other ones, and in 
particular \eqref{eq:rhoe} provides better results.
As the number of datapoints 
get large enough, GMM starts to be as accurate as clustering
based on energy statistics, as it should since it is 
consistent model to the data. In Figure~\ref{fig:consist}b, however, we
use the same parameters as in \eqref{eq:20gauss} but now with data
log-normally distributed:
\begin{equation}
\label{eq:20loggauss}
x \sim \tfrac{1}{2}\left[ e^{\mathcal{N}(\mu_1,\Sigma_1)}
+ e^{\mathcal{N}(\mu_2, \Sigma_2)}\right].
\end{equation}
We see that clustering based on energy statistics still performs accurately
for this kind of data, while $k$-means works a little bit better than chance,
and GMM is not even able to estimate the parameters. Again, \eqref{eq:rhoe}
provides slightly better results than \eqref{eq:standard_metric} or
\eqref{eq:rhohalf}. Notice also that Algorithm~\ref{algo} performs better
than Algorithm~\ref{kmeans_algo}. Both experiments in Figure~\ref{fig:consist}
shows that energy statistics clustering is nonparametric, since it is able
to cluster data coming from very different distributions.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss.pdf}\\[-1.0em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{loggauss.pdf}\\[-1.0em]
(b)
\end{minipage}
\caption{
\label{fig:consist}
(a) Data normally distributed as in 
\eqref{eq:20gauss}. We increase the number of points in each cluster to
illustrate the statistical consistency of the algorithms.
(b) The same experiment but for data following \eqref{eq:20loggauss}.
In both experiments, for each case we run every algorithm $100$ times and show
the average results. One can see the better performance of energy statistics
clustering, Algorithm~\ref{algo}, and in particular by using the semimetric
\eqref{eq:rhoe}. These two figures illustrate that energy statistics clustering
is nonparametric since it works well for very different distributions.
}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conclusion}

In this paper we have considered clustering from the perspective of energy
statistics, which provides a nonparametric test for equality of distributions.
Based on this, we showed that the clustering problem reduces to a quadratically
constrained optimization problem (QCQP), 
as described in Proposition~\ref{th:qcqp2}.
Moreover, we showed that clustering based on energy statistics is equivalent
to kernel $k$-means optimization problem, once the kernel is fixed; see
Proposition~\ref{th:kernel_kmeans}. Our results imply that kernel $k$-means
approach to clustering is actually a consequence of energy statistics
theory, and thus place this method into a principled statistical basis.
As already known \cite{Dhillon}, this approach is related to spectral
clustering, and graph partitioning problems. Therefore, all these problems 
may be seen as arising naturally from energy statistics clustering.
It is important to mention that energy statistics clustering, as formulated
here, is valid for
arbitrary metric spaces of negative type, and makes no assumptions about
the distribution of the data. Moreover, it does not rely on the concept of a
cluster mean, even implicitly.

We also proposed Algorithm~\ref{algo} as an alternative to the well-known
kernel $k$-means algorithm (see Algorithm~\ref{kmeans_algo}), 
where both have the same time
complexity. The numerical results show that Algorithm~\ref{algo}
might provide better clustering accuracy and is more robust 
than kernel $k$-means algorithm. Since there exists a huge literature about
kernel $k$-means, and approximation methods to make it faster, with
applications to several artificial and real data, 
we limited ourselves to analyze few but carefully designed experiments, which
illustrates the advantages of Algorithm~\ref{algo}.

\bigskip

%\subsection*{Acknowledgements}
%We thank \ldots
